Homoiconicity Without Template Haskell

### 2016-06-08 paper/paper.tex
restructured paper; lots of changes to the heterogeneous functions for better pattern matching

%\title{What's the point of homoiconicity?}
\title{The (floating) point of homoiconicity}
\subtitle{Functional Pearl}

@@

\begin{abstract}
Floating point numbers are dangerous.
Due to their limited precision,
seemingly correct mathematical formulae can give wildly inaccurate results.
This pearl shows that Haskell's type system can mitigate this danger.

The main observation is that many numerical Haskell functions are \emph{homoiconic}.
That is, we can easily convert them into isomorphic abstract syntax trees (ASTs).
The construction does not use template Haskell, and so works at both compile and run time.
We will use these ASTs to solve two problems with floating point numbers:
\begin{enumerate}
\item
We will write higher order functions that automatically stabilize (or optimize!) mathematical formulae.
%This lets one programmer write naive floating point code;
%then an expert on numerical analysis can write code that automatically improves the naive code.
\item
We will specify type class laws for floating point arithmetic that can be automatically tested.
Existing test frameworks do not work on floating point numbers because floats do not obey many traditional algebraic laws like associativity.
\end{enumerate}
We demonstrate these techniques using GHC's built-in numeric class hierarchy,
but these techniques generalize to user defined class hierarchies as well.
\end{abstract}

@@

Imagine you're writing a machine learning library in Haskell.
%You decide to start by implementing logistic regression.
%The first step is to implement the log-logistic function
%Logistic regression is one of the most important machine learning algorithms,
%and to implement it, you'll need to implement the log-logistic function
To implement the classic logistic regression algorithm,
you begin by implementing the log-logistic function
\begin{equation}
f(x) = \log\left(\frac{1}{1+\exp{(-x)}}\right)
\label{eq:log-logistic}
\end{equation}
%This function is used in logistic regression and many families of neural networks.
As a first attempt, you might write:
\begin{lstlisting}
logLogistic1 :: Floating x => x -> x
logLogistic1 x = log(1/(1+exp(-x)))
\end{lstlisting}
Unfortunately, this straightforward implementation is numerically unstable.
When specialized to the |Double| type,
the |exp| function overflows to infinity for inputs greater than (approximately) $709$.
%This overflow causes |logLogistic| to take the |log| of 0, which is undefined.
Therefore, on an input of $-710$,
|logLogisitic1| will take the |log| of 0,
returning negative infinity when the true answer is actually very close to $-710$.
Such an inaccurate answer is unacceptable,
so our function must be rewritten.

With some careful algebra, we can rewrite our log-logistic function in the following numerically stable form.
\begin{lstlisting}
logLogistic2 :: (Floating x, Ord x) => x -> x
logLogistic2 x = m+log(1/(exp(m)+exp(-x+m)))
    where
        m = min 0 x
\end{lstlisting}
It is hardly obvious that these two functions are meant to calculate the same expression.
The second implementation not only contains many more terms,
it has a different type signature due to the use of the |min| function.

Constructing these numerically stable functions is a notoriously difficult black art.\footnote{
For some classic introductions, see \cite{goldberg1991,higham2002,hamming2012}.
}
Wouldn't it be nice to have a tool to automate this procedure?
We'll construct a simple one in Section 3.

The logistic regression algorithm operates on vector spaces;
so you're next task is to get a proper vector space implementation.
Since you plan to use the foreign function interface heavily,
you decide to represent vectors as a pointer to a memory location.
\begin{lstlisting}
newtype Vector (dim::Nat) a = Vector (ForeignPtr a)
\end{lstlisting}
The |dim| type parameter encodes the dimension of the vector space.
It makes sense to store this information at the type level because vector space operations don't change a vector's dimension.

To make our |Vector| type useful,
we must define how to perform the standard mathematical operations on it.
You begin by making it an instance of |Num|.
Since you're a careful programmer, you decide to write test cases to ensure your |Num| instance is correct.
You begin by writing a QuickCheck property to ensure that your |(+)| function is properly associative.
It works for |Int|s, but misteriously fails on |Float|s and |Double|s.
%Unfortunately, the standard Haskell toolkit is not sufficient for the challenge.

\newpage
%Fortunately, a simple technique can be used to improve the stability of the log-logistic function.
%First, check our input to see if it would overflow the |exp| function.
%If no overflow would occur, then calculate our function as normal;
%but if overflow would occur, then use an approximation that does not rely on a call to |exp|.
%Performing this transformation gives us:
%\begin{lstlisting}
%logLogistic2 :: (Floating x, Ord x) => x->x
%logLogistic2 x = if x < (-709)
    %then x
    %else log(1/1+exp(-x))
%\end{lstlisting}
%This new function is numerically stable, but we've introduced two new problems:

We've made our function numerically stable, but we've introduced two new problems:

\begin{problem}[obfuscation]
The purpose of the if statement in |logLogistic2| is non-obvious.
To understand its purpose,
the reader must carefully consider the limits of floating point arithmetic.
Anyone who doesn't regularly work with floating point numbers is unlikely to discover this purpose on their own.
By transforming our function to have the correct floating point semantics,
we've obfuscated the function's purpose and made it harder for others to understand and contribute to our library.
\end{problem}

%\begin{problem}[specialization]
%The \lstinline{exp} function when specialized to \lstinline{Float}s overflows on values greater than (approximately) 88.
%This means that |logLogistic2| will give incorrect answers for inputs between 88 and 709.
%To properly handle both |Float|s and |Double|s,
%we must write two separate functions.
%In practice, we may want to support many other numeric types as well.
%Half precision (16 bit) floats and fixed point arithmetic are becoming increasingly popular in machine learning applications due to their increased speed, smaller memory footprint, and lower energy consumption \cite{gupta2015}.
%We need to manually reimplement the log-logistic function specialized to every numeric type we want our library to support.
%\end{problem}

These problems are not unique to the log-logistic function.
All functions over the real or complex numbers need to be transformed for floating point stability,
and all such manual transformations suffer from these problems.
This pearl presents a method for alleviating these problems.
Specifically, we will see how to write a function
%\begin{lstlisting}
%stabilize :: (Floating x, Ord x)
    %=> (x -> x) -> (x -> x)
%\end{lstlisting}
\begin{lstlisting}
stabilize :: (Floating x,Ord x) => (x->x) -> (x->x)
\end{lstlisting}
that automatically takes an input function like |logLogistic1| and outputs a numerically stable version of that function.
Unfortunately, this introduces a third problem:

\begin{problem}[unlawfulness]
%How do we know what transformations are valid?
The standard way to encode valid program transformations in Haskell is through type class laws.
A classic example is the |Semigroup| type class
\begin{lstlisting}
class Semigroup a where
    (+) :: a -> a -> a
\end{lstlisting}
which has the law that |(+)| should be associative.
%That is, for every type |a| that is a valid instance of |Semigroup|,
%|(a1+a2)+a3::a| should be equal to |a1+(a2+a3)::a|.
That is, if the type |a| is an instance of |Semigroup|,
and the variables |a1|, |a2|, and |a3| all have type |a|,
then |(a1+a2)+a3| should be equal to |a1+(a2+a3)|.
This is not true of floating point numbers, however.
The expression |(0.1+0.2)+0.3| is famously not equal to |0.1+(0.2+0.3)| for |Double|s.
But there is clearly a sense in which |Double|s are approximately associative,
and performing an associativity transformation cannot change our answer ``too much.''
%How can we formalize this notion?
In order for our |stabilize| method to be generic and useful,
we need a way to formalize this notion.
\end{problem}

This pearl attempts to solve all three problems.
First, we provide a mechanical procedure for constructing FAlgebras from what we call \emph{homogeneous} type classes.
(Both terms will be defined formally in the next section.)
These FAlgebras make our numeric functions \emph{homoiconic};
that is, we can easily get a syntax tree representing the function.
Second, we show how these syntax trees let us easily manipulate our numeric functions to implement the |stabilize| function described above.
Third, we introduce a notion of \emph{topological class laws}.
These class laws let us formalize the intuition that floating point numbers approximately obey class laws.
Finally, we extend our construction of FAlgebras to the \emph{heterogeneous} case.
Heterogeneous type classes let us represent the complicated numeric type class hierarchies that arise in practice when working with numeric types.

\section{Homogeneous Functions are homoiconic}
Consider the simple numeric class hierarchy shown in Figure \ref{code:group}.
With this class hierarchy, we can write a function that measures the degree to which the |(+)| operation obeys the associative rule.
\begin{lstlisting}
associator :: Group a => a -> a -> a -> a
associator a1 a2 a3
    = ((a1+a2)+a3) - (a1+(a2+a3))
\end{lstlisting}
We call the |associator| function \emph{homogeneous} because:
\begin{enumerate}
\item
The function's return type and all of its parameters are the same type variable.
\item
The constraints contain exactly one type class, and
that type class has exactly one parameter.
\end{enumerate}
Notice that each of the class methods for |Semigroup|, |Monoid|, and |Group| are also homogeneous.
Therefore, we will also say these classes are homogeneous.

In this section, we will see how to convert any homogeneous function into an isomorphic abstract syntax tree (AST).
This ability is called \emph{homoiconicity},
and is usually associated with Lisps.
We will then see several methods for evaluating these ASTs.
In subsequent sections, we will manipulate these ASTs to make floating point operations more convenient.

\begin{figure}
\begin{lstlisting}
class Semigroup a where
    (+) :: a -> a -> a

class Semigroup a => Monoid a where
    zero :: a

class Monoid a => Group a where
    (-) :: a -> a -> a
    negate :: a -> a
\end{lstlisting}
\caption{A simple numeric hierarchy that obeys the restrictions for homogeneous FAlgebras.}
\label{code:group}
\end{figure}

%\begin{figure}
%\begin{lstlisting}
%class Num a where
    %(+), (-), (*)       :: a -> a -> a
    %negate              :: a -> a
    %abs, signum         :: a -> a
    %fromInteger         :: Integer -> a
%
%class Num a => Fractional a where
    %(/)                 :: a -> a -> a
    %recip               :: a -> a
    %fromRational        :: Rational -> a
%
%class Fractional a => Floating a where
    %pi                  :: a
    %exp, log, sqrt      :: a -> a
    %(**), logBase       :: a -> a -> a
    %sin, cos, tan       :: a -> a
    %asin, acos, atan    :: a -> a
    %sinh, cosh, tanh    :: a -> a
    %asinh, acosh, atanh :: a -> a
%
%class Eq a where
    %(==), (/=)          :: a -> a -> Bool
%
%class Eq a => Ord a where
    %compare         :: a -> a -> Ordering
    %(<),(<=),(>),(>=)   :: a -> a -> Bool
    %max, min            :: a -> a -> a
%\end{lstlisting}
%\caption{
    %A portion of the numeric and comparison hierarchies defined in GHC's Prelude.
    %Many class methods satisfy the requirements for homogeneous FAlgebras, except:
    %%|fromInteger| and |fromRational| have non-variable parameters;
    %%and |(==)|, |(/=)|, |(<)|, |(<=)|, |(>)|, and |(>=)| have non-variable return types.
%}
%\label{code:ghc}
%\end{figure}

\subsection{A type class for FAlgebras}
Our construction is closely related to other constructions involving initial algebras, FAlgebras, and free monads.
See for example \cite{swierstra2008}.
The heart of our construction is the |FAlgebra| type class, defined as
\begin{lstlisting}
class
    ( Functor (Sig alg)
    , Foldable (Sig alg)
    ) => FAlgebra (alg::Type->Constraint)
        where
    data Sig alg a
    runSig :: alg a => Sig alg a -> a
\end{lstlisting}
The |FAlgebra| class is unusual in that its parameter has kind |Type->Constraint|.
This means that instances of |FAlgebra| will be other type classes (instead of types or type constructors).
The || in |FAlgebra| stands for homogeneous.
Most of the names used in this section will be prefixed with || to differentiate them from related names used in the later heterogeneous constructions.

The easiest way to explain the |FAlgebra| class is by example.
We'll start with the |Semigroup| type class defined above.
\begin{lstlisting}
instance FAlgebra Semigroup where
    data Sig Semigroup a = Sig_plus a a
    runSig (Sig_plus a1 a2) = a1+a2
\end{lstlisting}
The |Sig| associated data family is meant to encode what mathematicians call the \emph{signature} of the FAlgebra.
That is, for every class method, |Sig| should have a constructor with the same number of parameters.
Since |Semigroup| has a single function that takes two parameters,
the |Sig Semigroup| data instance has a single constructor that takes two parameters.
The |runSig| function evaluates the signature.
For each constructor in |Sig|, |runSig| should apply the corresponding function to the constructor's parameters.

To complete the |FAlgebra Semigroup| instance, we must also define the following superclass instances.
\begin{lstlisting}
instance Functor (Sig Semigroup) where
    fmap f (Sig_plus a1 a2)
        = Sig_plus (f a1) (f a2)

instance Foldable (Sig Semigroup) where
    foldr f b (Sig_plus a1 a2)
        = f a2 (f a1 b)
\end{lstlisting}
As usual, there is only a single valid |Functor| instance.
There are multiple valid |Foldable| instances depending on the order in which |foldr| considers the constructors' parameters,
but for our purposes the order is not important.

Instances get slightly more complicated when the parameter class has superclasses.
In this case, the |Sig| data instance must also have a constructor for each superclass that stores the superclass's signature.
This can be seen in the |Monoid| and |Group| instances below.
%The |FAlgebra| instances of |Monoid| and |Group| demonstrate this.
\begin{lstlisting}
instance FAlgebra Monoid where
    data Sig Monoid a
        = Sig_zero
        | Sig_Semigroup_Monoid
            (Sig Semigroup a)
    runSig Sig_zero = zero
    runSig (Sig_Semigroup_Monoid s)
        = runSig s

instance FAlgebra Group where
    data Sig Group a
        = Sig_negate a
        | Sig_minus a a
        | Sig_Monoid
            (Sig_Monoid_Group a)
    runSig (Sig_negate a) = a
    runSig (Sig_minus a1 a2) = a1-a2
    runSig (Sig_Monoid_Group s)
        = runSig s
\end{lstlisting}
Notice that the data instance |Sig Group| does not need a constructor for the |Semigroup| signature because this is included in the |Monoid| signature.
The corresponding |Functor| and |Foldable| instances have been omitted for brevity,
but can be found in the paper's git repo.

%The procedure for generating |FAlgebra| instances is purely mechanical.
%The ?? library automates the process with the |mkHomFAlgebra| template Haskell function.

@@

We can convert the signature of an FAlgebra into an AST as follows:
\begin{lstlisting}
type AST alg a = Free (Sig alg) a

data Free f a
    = Pure a
    | Free (f (Free f a))
\end{lstlisting}
Intuitively,
the |Pure| constructor represents a leaf in the AST,
and the |Free| constructor represents a branch.
The parameter to |Free| is filled by the appropriate |Sig| data instance.
The constructor used for the |Sig| instance corresponds to an operation,
and the parameters to the constructor will contain ASTs nested recursively.

For example, consider the expression |(1+2)+3|.
We can create a corresponding AST using |Semigroup|'s |FAlgebra| instance.
\begin{lstlisting}
Free
  (Sig_plus
    (Free
      (Sig_plus
        (Pure 1)
        (Pure 2)
      )
    )
    (Pure 3)
  )
\end{lstlisting}
We can also create a corresponding AST for any class that has |Semigroup| as a superclass.
Below is the |Group| AST.
\begin{lstlisting}
Free
  (Sig_Monoid_Group
    (Sig_Semigroup_Monoid
      (Sig_plus
        (Free
          (Sig_Monoid_Group
            (Sig_Semigroup_Monoid
              (Sig_plus
                (Pure 1)
                (Pure 2)
              )
            )
          )
        )
        (Pure 3)
      )
    )
  )
\end{lstlisting}
Notice that the |Sig_plus| constructor must be embedded into the |Group|'s |Sig| with calls to |Sig_Monoid_Group| and |Sig_Semigroup_Monoid|.

Constructing these ASTs by hand is a tedious process.
There is a considerable amount of boilerplate in the |Group| AST above as a result of the need to embed the |Sig_plus| constructor.
We can avoid this boilerplate with the following type class.
\begin{lstlisting}
class
    ( FAlgebra alg1
    , FAlgebra alg2
    ) => View alg1 alg2
        where
    embedHomSig
        :: Sig alg1 a -> Sig alg2 a
    unsafeExtractHomSig
        :: Sig alg2 a -> Sig alg1 a
\end{lstlisting}
Instances of the |View| class should satisfy the property that |alg1| is either equal to |alg2| or is a superclass of |alg2|, and
all such relationships should have a corresponding instance.
For example, |Semigroup| is a superclass of |Group|,
so we should have a |View Semigroup Group| instance.
\begin{lstlisting}
instance View Semigroup Group
    embedHomSig s
        = Sig_Monoid_Group (embedHomSig s)
    unsafeExtractHomSig (Sig_Monoid_Group s)
        = unsafeExtractHomSig s
\end{lstlisting}
Because of this property, we know that the |Sig alg1 a| data instance can always be embedded into a |Sig alg2 a| data instance.
But it is only sometimes true that we can extract a |Sig alg1 a|,
which is why the corresponding function is labeled |unsafe|.

Armed with the |View| class, we can create a uniform representation for all class's ASTs.
The expression |(1+2)+3| can now be represented as:
\begin{lstlisting}
Free
  (embedHomSig
    (Sig_plus
      (Free
        (embedHomSig
          (Sig_plus
            (Pure 1)
            (Pure 2)
          )
        )
      )
      (Pure 3)
    )
  )
\end{lstlisting}

%Every FAlgebra can be embedded into itself.
%\begin{lstlisting}
%instance View alg alg where
    %embedSig = id
    %unsafeExtractSig = id
%\end{lstlisting}

%\begin{lstlisting}
%instance View Semigroup Monoid where
    %embedSig = Sig_Monoid_Semigroup
    %unsafeExtractSig (Sig_Monoid_Semigroup s) = s

%instance View Monoid Group where
    %embedSig s = Sig_Group_Monoid (embedSig s)
    %unsafeExtractSig (Sig_Group_Monoid s) = unsafeExtractSig s
%\end{lstlisting}

%\begin{lstlisting}
%instance View Semigroup Group where
    %embedSig s = Sig_Group_Monoid (embedSig s)
    %unsafeExtractSig (Sig_Group_Monoid s) = unsafeExtractSig s
%\end{lstlisting}

Using our uniform representation of an AST,
we can make the AST itself an instance of any homogeneous type class.
Again, we'll approach the procedure by example.
We can write a |Semigroup| instance for |AST Semigroup a| as:
\begin{lstlisting}
instance
    ( View Semigroup alg
    ) => Semigroup (Free (Sig alg) a)
        where
    (+) e1 e2 = Free $ embedHomSig $
            Sig_plus e1 e2
\end{lstlisting}
For each function in the type class,
we simply embed the function's constructor into the AST by prepending the expression |Free $ embedHomSig $|.
The |Monoid| and |Group| instances follow a similar pattern.
\begin{lstlisting}
instance
    ( View Semigroup alg
    , View Monoid alg
    ) => Monoid (Free (Sig alg) a)
        where
    zero = Free $ embedHomSig $
        Sig_zero

instance
    ( View Semigroup alg
    , View Monoid alg
    , View Group alg
    ) => Group (Free (Sig alg) a)
        where
    negate e  = Free $ embedHomSig $
        Sig_negate e
    (-) e1 e2 = Free $ embedHomSig $
        Sig_minus e1 e2
\end{lstlisting}
We can easily create ASTs using these instances.
The |Semigroup| AST can be created with the expression
\begin{lstlisting}
(Pure 1+Pure 2)+Pure 3
    :: AST Semigroup a
\end{lstlisting}
And the |Group| AST can be created by just changing the type signature.
\begin{lstlisting}
(Pure 1+Pure 2)+Pure 3
    :: AST Group a
\end{lstlisting}

@@

The are no tricks in the |Show| instance for |Semigroup|'s |Sig|.
\begin{lstlisting}
instance Show a =>
    Show (Sig Semigroup a)
        where
    show (Sig_plus a1 a2)
        = show a1++"+"++show a2
\end{lstlisting}
For each operation constructor, we simply show the parameters and the operation connecting them.
For type classes with superclasses, we also need to recursively show the superclasses as can be seen in the |Monoid| and |Group| instances below.
\begin{lstlisting}
instance Show a =>
    Show (Sig Monoid a)
        where
    show (Sig_Semigroup_Monoid s)
        = show s
    show (Sig_zero) = "zero"

instance Show a =>
    Show (Sig Group a)
        where
    show (Sig_Monoid_Group s)
        = show s
    show (Sig_negate a)
        = "negate "++show a
    show (Sig_minus a1 a2)
        = show a1++"-"++show a2
\end{lstlisting}
The |Free| type's |Show| instance is similarly straightforward..
\begin{lstlisting}
instance (Show a, Show (f (Free f a)))
    => Show (Free f a)
        where
    show (Pure a) = show a
    show (Free f) = "("++show f++")"
\end{lstlisting}
If the syntax tree contains only a single leaf, we just show that leaf.
If the syntax tree contains a branch,
then put parenthesis around the branch so that we don't lose information about the order of operations, and show the functor in the middle.

>>>>

%\title{What's the point of homoiconicity?}
%\title{The (floating) point of homoiconicity}
\title{Homoiconicity Without Template Haskell}
\subtitle{Functional Pearl}

@@

\begin{abstract}
We say a function is \emph{homoiconic} if it can be converted into an isomorphic abstract syntax tree (AST),
and the resulting AST can be converted back into the original function.
Homoiconicity is a useful property because it lets libraries implement arbitrary program transformations on these functions by manipulating the AST.
This pearl shows that many Haskell functions are homoiconic.
The technique does not rely on template Haskell,
and so works on expressions available at either compile or run time.
To motivate this homoiconicity,
we construct a small library that automatically transforms numerically unstable code into stable code.

%This pearl shows that many Haskell functions are \emph{homoiconic}.
%That is, we can convert these functions into isomorphic abstract syntax trees (ASTs),
%and these ASTs can be converted back into their original Haskell functions.
%%These ASTs are ordinary Haskell data types,
%%so they can be manipulated arbitrarily to perform program transformations.
%The construction does not require template Haskell,
%and so works on expressions available at either compile or run time.
%To motivate this homoiconicity,
%we construct a small library that automatically transforms numerically unstable code into stable code.

%Floating point numbers are dangerous.
%Due to their limited precision,
%seemingly correct mathematical formulae can give wildly inaccurate results.
%This pearl shows that Haskell's type system can mitigate this danger.
%
%The main observation is that many numerical Haskell functions are \emph{homoiconic}.
%That is, we can easily convert them into isomorphic abstract syntax trees (ASTs).
%The construction does not use template Haskell, and so works at both compile and run time.
%We will use these ASTs to solve two problems with floating point numbers:
%\begin{enumerate}
%\item
%We will write higher order functions that automatically stabilize (or optimize!) mathematical formulae.
%%This lets one programmer write naive floating point code;
%%then an expert on numerical analysis can write code that automatically improves the naive code.
%\item
%We will specify type class laws for floating point arithmetic that can be automatically tested.
%Existing test frameworks do not work on floating point numbers because floats do not obey many traditional algebraic laws like associativity.
%\end{enumerate}
%We demonstrate these techniques using GHC's built-in numeric class hierarchy,
%but these techniques generalize to user defined class hierarchies as well.
\end{abstract}

@@

Imagine you're writing a machine learning library in Haskell.
%You decide to start by implementing logistic regression.
%The first step is to implement the log-logistic function
%Logistic regression is one of the most important machine learning algorithms,
%and to implement it, you'll need to implement the log-logistic function
%To implement the classic logistic regression algorithm,
%you begin by implementing the log-logistic function
At some point, you'll need to implement the log-logistic function
\begin{equation}
f(x) = \log\left(\frac{1}{1+\exp{(-x)}}\right)
\label{eq:log-logistic}
\end{equation}
This function is used in logistic regression and many families of neural networks.
As a first attempt, you might write:
\begin{lstlisting}
logLogistic1 :: Floating x => x -> x
logLogistic1 x = log(1/(1+exp(-x)))
\end{lstlisting}
Unfortunately, this straightforward implementation is numerically unstable.
When specialized to the |Double| type,
the |exp| function overflows to infinity for inputs greater than (approximately) $709$.
%This overflow causes |logLogistic| to take the |log| of 0, which is undefined.
Therefore, on an input of $-710$,
|logLogisitic| will take the |log| of 0,
returning negative infinity when the true answer is actually very close to $-710$.
Such an inaccurate answer is unacceptable,
so our function must be rewritten.

But like all good programmers, you're lazy.
Your machine learning library will contain thousands of numeric functions,
all of which will need to be carefully written to avoid numeric instabilities.
What a maintenance nightmare!
Soon you start daydreaming about a function
\begin{lstlisting}
stabilize :: Floating a => (a -> a) -> (a -> a)
\end{lstlisting}
that takes a naively written floating point function as input and outputs a numerically stable version.
Armed with such a function, you could write a stable implementation of the log-logistic function as
\begin{lstlisting}
logLogistic2 :: Floating x => x -> x
logLogistic2 x = stabilize $ log(1/(1+exp(-x)))
\end{lstlisting}
%stopzone

In this pearl, we see how to write functions like |stabilize|.
The key observation is that certain Haskell functions can be converted into an isomorphic abstract syntax tree (AST).
The |stabilize| function uses this fact to convert its input function into an AST;
manipulates that AST to remove numerical stabilities;
then converts the AST back into a callable function.

In Section \ref{sec:homogeneous}, we introduce the idea of a \emph{homogeneous} function and show that these functions are homoiconic.
In Section \ref{sec:stabilize}, we use the construction above to implement several versions of the |stabilize| function.
In Section \ref{sec:heterogeneous}, we introduce the idea of \emph{heterogeneous} functions and show that they are also homoiconic.
Finally, we conclude by introducing a Haskell package called {\ttfamily homoiconic} that makes it easy for you to .

\section{Homogeneous functions are homoiconic}
\label{sec:homogeneous}
We call a function \emph{homogeneous} if:
\begin{enumerate}
\item
%The function's return type and all of its parameters are the same type variable.
The function is polymorphic in its return type.
\item
The function's parameters are either the same as the return type or they are concrete.
\item
The constraints contain exactly one type class, and
that type class has exactly one parameter,
which is the function's return type.
\end{enumerate}
The |logLogistic1| function is homogeneous.
Figure \ref{code:ghc} shows many more examples from GHC's numeric and ordering hierarchies.

In this section, we will show that every type class has an associated monad representing the abstract syntax tree of the class's homogeneous functions.
Our construction is closely related to other constructions involving initial algebras, FAlgebras, and free monads.
See for example \cite{swierstra2008}.

\begin{figure}
\begin{lstlisting}
class Num a where
    (+), (-), (*)       :: a -> a -> a
    negate              :: a -> a
    abs, signum         :: a -> a
    fromInteger         :: Integer -> a

class Num a => Fractional a where
    (/)                 :: a -> a -> a
    recip               :: a -> a
    fromRational        :: Rational -> a

class Fractional a => Floating a where
    pi                  :: a
    exp, log, sqrt      :: a -> a
    (**), logBase       :: a -> a -> a
    sin, cos, tan       :: a -> a
    asin, acos, atan    :: a -> a
    sinh, cosh, tanh    :: a -> a
    asinh, acosh, atanh :: a -> a

class Eq a where
    (==), (/=)          :: a -> a -> Bool

class Eq a => Ord a where
    compare             :: a -> a -> Ordering
    (<),(<=),(>),(>=)   :: a -> a -> Bool
    max, min            :: a -> a -> a
\end{lstlisting}
\caption{
    A portion of the numeric and comparison hierarchies defined in GHC's Prelude.
    %Functions highlighted in bold are homogeneous,
    %and the remainder are not.
}
\label{code:ghc}
\end{figure}

\subsection{A type class for FAlgebras}
FAlgebras are a technique used in the mathematical fields of universal algebra and category theory for defining algebraic structures.
We represent an FAlgebra with the following type class.
\begin{lstlisting}
class Functor (Sig alg) => FAlgebra alg where
    data Sig alg a
    runSig :: alg a => Sig alg a -> a
\end{lstlisting}
The |FAlgebra| class is unusual in that its parameter has kind |Type->Constraint|.
This means that instances of |FAlgebra| will be other type classes
(instead of types or type constructors).

It is easiest to understand instances of |FAlgebra| by walking through an example.
Below is the instance for |Fractional|.
\begin{lstlisting}
instance FAlgebra Fractional where
    data Sig Fractional a
        = Sig_div a a
        | Sig_recip a
        | Sig_fromRational Rational
        | Sig_Fractional_Num (Sig Num a)
    runSig (Sig_div a1 a2)        = a1/a2
    runSig (Sig_recip a)          = recip a
    runSig (Sig_fromRational r)   = fromRational r
    runSig (Sig_Fractional_Num s) = runSig s
\end{lstlisting}

The data family |Sig| encodes what mathematicians call the \emph{signature} of the FAlgebra.
The signature defines all the operations that can be performed on an FAlgebra.
Each |Sig| data instance can have many constructors,
which can be divided into two categories.
First, there should be one constructor for each homogeneous class function.
This constructor should have the same parameters as the function.
For the |Fractional| example, we've defined the three constructors |Sig_div|, |Sig_recip|, and |Sig_fromRational| for the three member functions |(/)|, |recip|, and |fromRational|.
Second, there should be a constructor for each superclass.
This constructor should store the |Sig| of the corresponding superclass.
For the |Fractional| example, we have the |Sig_Fractional_Num| constructor corresponding to the |Num| superclass.

The |runSig| class method gives us a way to evaluate any |Sig|.
For each function constructor, |runSig| should call the corresponding function.
For each superclass constructor, |runSig| recursively calls |runSig| on the superclass's |Sig|.

The F in FAlgebra comes from the fact that every signature is actually a functor.
In Haskell, we encode this by enforcing that |Sig alg| must be an instance of the |Functor| class in order to create a valid |FAlgebra| instance for |alg|.
As usual, there is only a single valid |Functor| instance.
For |Fractional|, it is:

\begin{lstlisting}
instance Functor (Sig Fractional) where
    fmap f (Sig_div a1 a2) = Sig_div (f a1) (f a2)
    fmap f (Sig_recip a) = Sig_recip (f a)
    fmap f (Sig_fromRational r) = Sig_fromRational r
    fmap f (Sig_Fractional_Num s) = fmap f s
\end{lstlisting}
As usual, there is only a single valid |Functor| instance.
There are multiple valid |Foldable| instances depending on the order in which |foldr| considers the constructors' parameters,
but for our purposes the order is not important.

@@

For every FAlgebra, there is an associated AST with the corresponding algebraic structure.
This AST is sometimes called an \emph{initial algebra},
and has a standard construction via the \emph{free monad}.

In Haskell, the free monad is defined as
\begin{lstlisting}
data Free f a
    = Pure a
    | Free (f (Free f a))
\end{lstlisting}
which lets us define our AST as
\begin{lstlisting}
type AST alg a = Free (Sig alg) a
\end{lstlisting}
Intuitively,
the |Pure| constructor represents a leaf in the AST,
and the |Free| constructor represents a branch.
The parameter to |Free| is filled by the appropriate |Sig| data instance.
The constructor used for the |Sig| instance corresponds to an operation,
and the parameters to the constructor will contain ASTs nested recursively.

For example, consider the expression |(1+2)+3|.
We can create a corresponding AST using |Num|'s |FAlgebra| instance.
\begin{lstlisting}
Free
  (Sig_plus
    (Free
      (Sig_plus
        (Pure 1)
        (Pure 2)
      )
    )
    (Pure 3)
  )
\end{lstlisting}
We can also create a corresponding AST for any class that has |Num| as a superclass.
Below is the same expression encoded using |Floating|'s |FAlgebra| instance.
\begin{lstlisting}
Free
  (Sig_Floating_Fractional
    (Sig_Fractional_Num
      (Sig_plus
        (Free
          (Sig_Floating_Fractional
            (Sig_Fractional_Num
              (Sig_plus
                (Pure 1)
                (Pure 2)
              )
            )
          )
        )
        (Pure 3)
      )
    )
  )
\end{lstlisting}
Notice that the |Sig_plus| constructor must be embedded into |Floating|'s |Sig| with calls to |Sig_Floating_Fractional| and |Sig_Fractional_Num|.

Constructing these ASTs by hand is a tedious process.
There is a considerable amount of boilerplate in the |Floating| AST above as a result of the need to embed the |Sig_plus| constructor.
We can avoid this boilerplate with the following type class.
\begin{lstlisting}
class
    ( FAlgebra alg1
    , FAlgebra alg2
    ) => View alg1 alg2
        where
    embedSig         :: Sig alg1 a -> Sig alg2 a
    unsafeExtractSig :: Sig alg2 a -> Sig alg1 a
\end{lstlisting}
Instances of the |View| class should satisfy the property that |alg1| is either equal to |alg2| or is a superclass of |alg2|, and
all such relationships should have a corresponding instance.
For example, |Num| is a superclass of |Floating|,
so we should have a |View Num Floating| instance.
\begin{lstlisting}
instance View Num Floating
    embedSig s
        = Sig_Fractional_Floating (embedSig s)
    unsafeExtractSig (Sig_Fractional_Floating s)
        = unsafeExtractSig s
\end{lstlisting}
Because of this property, we know that the |Sig alg1 a| data instance can always be embedded into a |Sig alg2 a| data instance.
But it is only sometimes true that we can extract a |Sig alg1 a|,
which is why the corresponding function is labeled |unsafe|.

Armed with the |View| class, we can create a uniform representation for all class's ASTs.
The expression |(1+2)+3| can now be represented as:
\begin{lstlisting}
Free
  (embedSig
    (Sig_plus
      (Free
        (embedSig
          (Sig_plus
            (Pure 1)
            (Pure 2)
          )
        )
      )
      (Pure 3)
    )
  )
\end{lstlisting}

Using our uniform representation of an AST,
we can make the AST itself an instance of any |FAlgebra| type class.
Again, we'll approach the procedure by example.
We can write a |Num| instance for |AST Num a| as:
\begin{lstlisting}
instance View Num alg => Num (Free (Sig alg) a)
        where
    (+) e1 e2 = Free (embedSig (Sig_plus e1 e2))
\end{lstlisting}
For each function in the type class,
we simply embed the function's constructor into the AST by prepending the expression |Free $ embedSig $|.
The |Fractional| and |Floating| instances follow a similar pattern.

We can easily create ASTs using these instances.
The |Num| AST can be created with the expression
\begin{lstlisting}
(1+2)+3 :: AST Num a
\end{lstlisting}
And the |Floating| AST can be created by just changing the type signature.
\begin{lstlisting}
(1+2)+3 :: AST Floating a
\end{lstlisting}

@@

The are no tricks in the |Show| instance for the |Sig| instance.
The instance for |Fractional| is shown below.
\begin{lstlisting}
instance Show a => Show (Sig Fractional a) where
    show (Sig_div a1 a2) = show a1++"/"++show a2
    show (Sig_recip a) = "recip "++show a
    show (Sig_fromRational r) = "fromRational "++show r
    show (Sig_Fractional_Num s) = show s
\end{lstlisting}
For each function constructor, we simply show the corresponding function and its parameters.
If the function is an operator, we'll display it infix for convenience.
For each superclass constructor, we simply recursively call show on the superclass's |Sig|.

The |Free| type's |Show| instance is similarly straightforward.
\begin{lstlisting}
instance (Show a, Show (f (Free f a)))
    => Show (Free f a)
        where
    show (Pure a) = show a
    show (Free f) = "("++show f++")"
\end{lstlisting}
If the syntax tree contains only a single leaf, we just show that leaf.
If the syntax tree contains a branch,
then put parenthesis around the branch so that we don't lose information about the order of operations, and show the functor in the middle.

Now, when we type an expression into ghci,
\begin{lstlisting}
ghci> (1+2)+3
6
ghci> (1+2)+3 :: AST Floating Int
(1+2)+3
\end{lstlisting}

\subsection{Evaluating the syntax tree}
Another useful task for ASTs is to evaluate them,
generating the value they represent.
To do this, we will take advantage of the free monad's structure.
The free monad is so called because as long as the parameter |f| is a |Functor|,
then |Free| has a valid |Monad| instance.
Since all |Monad|s are by necessity also |Functor|s,
|Free| has a valid |Functor| instance as well.
\begin{lstlisting}
instance Functor f => Functor (Free f) where
    fmap g (Pure a) = Pure (g a)
    fmap g (Free f) = Free (fmap (fmap g) f)
\end{lstlisting}
For our purposes, that's all the structure we'll need.
